---
title: "Approximate Matching"
author: "Mark van der Loo and Edwin de Jonge"
date: "Tokyo | Institute of Statistical Mathemathics | 2020"
output: 
  beamer_presentation:
    include:
      in_header: ../00tex/presheader.tex
classoption: "aspectratio=169"
---

## Try the code together with your neighbour

\centering\huge
`02input/raw_to_input.R`


## String distance

### Default (Optimal String Alignment distance)

Count number of character deletions, insersions, substitutions and 
transpositions (of adjacent characters)
```{r}
library(stringdist)
stringdist("Ross Ihaka", "Robert Gentleman")
```


## Exact Matching with `match`

```{r}
lookup <- c("Alice","Bob","Carol","Danny")
raw    <- c("Bob","Carl","Rob","bob","Dan","Alice")
i <- match(raw, lookup)
data.frame(raw=raw, matched=lookup[i])
```

## Approximate Matching with `stringdist::amatch`

```{r}
library(stringdist)
j <- amatch(raw, lookup, maxDist=2)
data.frame(raw=raw, matched=lookup[i], amatched=lookup[j])
```

$\to$ Match with closest match, and distance $\leq 2$.

## Optimal string alignment?

```{r}
stringdist("Robert Gentleman", "Gentleman, Robert")
stringdist("Robert Gentleman", "Ross Ihaka")
```

$\to$ OSA wil give a false match (if we allow `maxDist` of 12)


## Alternative: cosine distance 

```{r}
stringdist("Robert Gentleman","Gentleman, Robert"
            , method="cosine", q=2)
stringdist("Robert Gentleman","Ross Ihaka"
            , method="cosine", q=2)
```

### Notes
- Based on counting co-occurrence of character $q$-grams (here: pairs).
- Always between 0 and 1



## More on `amatch`

```{r, eval=FALSE}
amatch(x, table, method, maxDist,...)
```

\begin{tabular}{ll}
\texttt{x}      & character data to be matched\\
\texttt{table}  & the lookup table with clean values\\
\texttt{method} & string distance type\\
\texttt{maxDist}& Maximum distance allowed (depends on \texttt{"method"}!)\\
\texttt{...}    & Extra options depending on \texttt{"method"}
\end{tabular}

### Example

```{r,eval=FALSE}
amatch(raw, lookup, method="cosine", maxDist=0.5, q=3)
```



## Assignment

Merge data from the `companies` dataset with data from `backbone.csv`.

- Using approximate matching on the `"name"` and `"company"` column.
- Think about and try different distance functions and `maxDist`
- Keep your best solution
- Remove rows that cannot be matched
- Write to `02input/myinput.csv`


# More on String distances

## More on string distances

### Main idea

Define a sence of _distance_ between two text strings

### Distance

A function $d(s,t)$ that takes two arguments and

- returns a nonnegative number,
- returns zero if and only if $s=t$
- is symmetric: $d(s,t)=d(t,s)$
- is the length of a shortest path between $s$ and $t$: $d(s,t)\leq d(s,u)+d(u,t)$

### Note

Some string distances violate one or more of the above assumptions.


## Distance types

- Edit based
- $q$-gram based
- Heuristic


## Edit-based string distances

### Idea

1. Choose basic steps to alter a string
2. Find the smallest nr of steps that changes $s$ into $t$
3. The distance equals the nr of steps needed.

### Basic steps

- deletion: hihi $\to$ hii
- insertion: hihi $\to$ hihih
- substitution: hihi $\to$ hiha
- transposition: hihi $\to$ ihhi


## Edit-based string distances

\begin{center}

  \begin{tabular}{p{2cm}cccc}
      &\multicolumn{4}{c}{\textbf{Allowed operation}}\\
      \textbf{Distance}  & substitution & deletion & insertion & transposition\\
      Hamming            & \yep{}   & \nope{} & \nope{} &\nope{} \\
      LCS                & \nope{}  & \yep{}  & \yep{}  &\nope{} \\
      Levenshtein        & \yep{}   & \yep{}  & \yep{}  &\nope{} \\
      OSA                & \yep{}   & \yep{}  & \yep{}  &\yep{}$^*$ \\
      Damerau-Levenshtein& \yep{}   & \yep{}  & \yep{}  &\yep{}\\
    \end{tabular}\\
  $^*$Substrings may be edited only once.
\end{center}

## Example

### Levenshtein distance

$$
leia\xrightarrow[+1]{\textrm{sub}}lela\xrightarrow[+1]{\textrm{ins}}leela
$$

### Longest common subsequence distance

$$
leia
\xrightarrow[+1]{\textrm{del}}lea
\xrightarrow[+1]{\textrm{ins}}leea
\xrightarrow[+1]{\textrm{ins}}leela
$$


## $q$-gram based distances (I)

### Algorithm

 - Tabulate substrings of length $q$ ($=$ $q$-gram profile)
 - Compute a distance between the profiles

### Example

2-gram profile of banana

\begin{tabular}{ccc}
ba&an&na\\
1 & 1& 2 
\end{tabular}

## $q$-gram based distances (II)

### $q$-gram distance

Manhattan distance between $q$-gram profiles
$$
\sum_{\sf{qgram}} |n_{\textsf{qgram}}(s) -n_{\textsf{qgram}}(t)|
$$

### Cosine-distance

1 minus the cosine of the angle between the profiles

$$
1 - \frac{\mathbf{n}(s)\cdot\mathbf{n}(t)}
{\|\mathbf{n}(s)\|\|\mathbf{n}(t)\|}
$$

### Note

 - Does not satisfy the 'identity' demand.
 - Often one chooses $q=2$ of $q=3$ 


## Jaro-Winkler distance

### Jaro distance
$$
d_j(s,t)=1-\frac{1}{3}\left(\frac{m}{|s|} + \frac{m}{|t|} + \frac{m + T}{m}\right)
$$

- $m$ number of matching characters (within a window)
- $T$ number of matches that need swapping
- $|s|,|t|$ number of characters in $s$, $t$.

### Jaro-Winkler distance
$$
d_{jw}(s,t) = [1-p\ell(s,t)]d_j(s,t)
$$

- $\ell(s,t)$ length of longest equal prefix (up to 4 characters)
- $p$ a number between $0$ and $0.25$ (usually $0.1$)

## Soundex

### Algorithm

- Strings are appointed a code: same code means 'sounds the same'
- Equal codes: distance zero, otherwise 1

### Example

- Farnsworth $\to$ H652
- Fnarswort $\to$  H562

### Note

- Based on English pronunciation
- Many extensions exist (see the `phonics` R package)


## Which one to use?

### Considerations

- Fixed versus variable structure/length
- Why would strings differ? (typos, speech, deliberate changes)
- Performance







